The Structure of Human Language Reflects Limitations of Working Memory

### 2019-04-25 paper-short.tex
revising

In order to test this hypothesis, it is necessary to consider an individual listener's language comprehension process.
%We consider language understanding from the perspective of a listener processing a stream of words uttered by an interlocutor.
During comprehension, this listener is processing a stream of words uttered by an interlocutor.
This listener will represent information about past words in their short-term memory.
We quantify the amount of memory resources that are used by the number of bits.
We consider the question of how many bits of information this listener has to carry from one instant to the next.  %\jd{why?}

Human language is well known to have a large amount of context-dependence in the meanings and forms of words.
Understanding what a word means often requires consideration of context. 
Consider the sentence `The climber gave the attempt to reach the peak up.'
Understanding the word \emph{up} requires establishing that it is related to the word \emph{gave} -- neither of them has the meaning of `give up' by itself.
In order to understand this sentence correctly, a listener has to store some information about the occurrence of the word \emph{gave} until reaching \emph{up}.
For language comprehension, this means that a listener's ease of comprehension will be modulated by how much memory they invest into representing information about past context:
A listener who remembers more of the past context should have less difficulty understanding, at the cost of increased working memory load. 
We thus expect there to be a tradeoff between a listener's ease of language understanding and their working memory load.

%We use information theory to quantify the tradeoff between memory load and ease of understanding.
Experimental research has established that human processing effort on a given word is strongly predicted by its surprisal, the degree to which it cannot be predicted from the preceding context:
%Experimental research has firmly established Surprisal as an accurate metric of human processing effort in language comprehesion.
%The surprisal of a word in context is the degree to which it cannot be predicted:
\begin{equation}\label{eq:surp}
	-\log p(x_i|x_{1...i-1})
\end{equation}
where $x_i$ is the $i$-th word in a given utterance (or discourse).
%This quantity is known to be a strong linear predictor of human processing effort.

A listener with limited memory of prior context will not be able to compute the full conditional distribution~(\ref{eq:surp}).
We denote a listener's memory representation at time $t$ by $L_t$, which is some -- generally imperfect -- summary of the preceding context $\dots x_{t-3}x_{t-2}x_{t-1}$.
The amount of information stored by the listener is the number of bits in $L_t$, i.e., its entropy $M := H[L_t]$.
This listener will experience processing effort
\begin{equation}
	S :=	-\log p(x_t|L_t)
\end{equation}
upon hearing the next word $x_t$.
For a listener who invests less bits into representing information $L_t$ about prior context, $L_t$ will be a more coarse-grained representation of past observations.
Thus, the predictions made by such a listener will be on average less accurate, and such a listener will incur greater processing effort $S$.
%Since $L_t$ is some summary representation, 
%Therefore, ne should expect that processing effort $S$ increases for a listener who invests less bits into representing information about prior context.\jd{why? this is important.}

%For instance, in `The climber gave the attempt to reach the peak up', the word \emph{up} is less surprising to a listener who has maintained some information about the word \emph{gave} in their memory.
Thus, there is a tradeoff between memory and surprisal:
A listener can achieve lower levels of processing effort, at the expense of increased memory load.

@@

%TODO
Let $I_t$ be the Conditional Mutual Information of two words $X_0, X_t$ conditioned on the intervening words:
\begin{equation}
	\operatorname{I}_t := \operatorname{I}[X_t, X_0 | X_1, \dots, X_{t-1}] = \operatorname{H}[X_t|X_1, \dots, X_{t-1}] - \operatorname{H}[X_t|X_0, \dots, X_{t-1}] 
\end{equation}
This quantity measures how much predictive information is provided by the next word $X_t$ by the word $t$ steps in the past.
It is a statistical property of the language.
Our theoretical results describes how the tradeoff between memory and comprehension difficulty relates to $I_t$.

@@

An immediate consequence of this theoretical result is a general relation between memory and locality:
A language will enable more efficient tradeoffs if $I_t$ decays faster as $t$ grows, which corresponds to stronger concentration of predictive information in the recent past.

@@

>>>>

\mhahn{smooth out}We consider an individual listener's language comprehension process, processing a stream of words uttered by an interlocutor.
Experimental research has established that listeners form probabilistic expectations about the upcoming words.
Consider example...
Human processing effort on a given word is strongly predicted by its surprisal, the degree to which it cannot be predicted from the preceding context.
Surprisal is known to be a strong linear predictor of processing effort as measured by per-word reading times.
For an idealized listener with unbounded memory resources, surprisal will be given by the negative log-probability of a word given its preceding context:
\begin{equation}\label{eq:surp}
	-\log p(x_i|x_{1...i-1})
\end{equation}
where $x_i$ is the $i$-th word in a given utterance (or discourse).

In reality, a listener will have bounded capacity to represent prior context.
A listener with limited memory of prior context will not be able to compute the full conditional distribution~(\ref{eq:surp}).
We denote a listener's memory representation at time $t$ by $L_t$, which is some -- generally imperfect -- summary of the preceding context $\dots x_{t-3}x_{t-2}x_{t-1}$.
This listener will experience surprisal
\begin{equation}\label{eq:surp-listener}
	S :=	-\log p(x_t|L_t)
\end{equation}
upon hearing the next word $x_t$.
%The amount of information stored by the listener is the number of bits in $L_t$, i.e., its entropy $M := H[L_t]$.
%For a listener who invests less bits into representing information $L_t$ about prior context, $L_t$ will be a more coarse-grained representation of past observations.
%Thus, the predictions made by such a listener will be on average less accurate, and such a listener will incur greater processing effort $S$.

Information theory, specifically the Data Processing Inequality, implies that the average surprisal $S$ will be higher for a listener who represents less information about past input.
Consequently, there is a tradeoff between memory and surprisal:
A listener can achieve lower levels of processing effort, at the expense of increased memory load.

Formally, we quantify memory as the number of bits about prior input that a listener represents.

@@

%TODO
Let $I_t$ be the Conditional Mutual Information of two words $X_0, X_t$ conditioned on the intervening words:
\begin{equation}
	\operatorname{I}_t := \operatorname{I}[X_t, X_0 | X_1, \dots, X_{t-1}] = \operatorname{H}[X_t|X_1, \dots, X_{t-1}] - \operatorname{H}[X_t|X_0, \dots, X_{t-1}] 
\end{equation}
This quantity measures how much predictive information is provided by the next word $X_t$ by the word $t$ steps in the past.
It is a statistical property of the language, and can be estimated from large-scale text data.
Our theoretical results describes how the tradeoff between memory and comprehension difficulty relates to $I_t$.

@@

An immediate consequence of this theoretical result is a general relation between memory and locality:
A language will enable more efficient tradeoffs if $I_t$ decays very fast as $t$ grows, which corresponds to strong concentration of predictive information in the recent past.

@@

-This listener will represent information about past words in their short-term memory.
We quantify the amount of memory resources that are used by the number of bits.
We consider the question of how many bits of information this listener has to carry from one instant to the next.  %\jd{why?}

-Human language is well known to have a large amount of context-dependence in the meanings and forms of words.
Understanding what a word means often requires consideration of context. 
Consider the sentence `The climber gave the attempt to reach the peak up.'
Understanding the word \emph{up} requires establishing that it is related to the word \emph{gave} -- neither of them has the meaning of `give up' by itself.
In order to understand this sentence correctly, a listener has to store some information about the occurrence of the word \emph{gave} until reaching \emph{up}.
For language comprehension, this means that a listener's ease of comprehension will be modulated by how much memory they invest into representing information about past context:
A listener who remembers more of the past context should have less difficulty understanding, at the cost of increased working memory load. 
We thus expect there to be a tradeoff between a listener's ease of language understanding and their working memory load.

%We use information theory to quantify the tradeoff between memory load and ease of understanding.
%This quantity is known to be a strong linear predictor of human processing effort.

%Since $L_t$ is some summary representation, 
%Therefore, ne should expect that processing effort $S$ increases for a listener who invests less bits into representing information about prior context.\jd{why? this is important.}

%For instance, in `The climber gave the attempt to reach the peak up', the word \emph{up} is less surprising to a listener who has maintained some information about the word \emph{gave} in their memory.
